---
# Code for: UAV-based high-throughput phenotyping of seedling vigor in mungbean
author: "Venkata Naresh Boddepalli"
date: "2023-12-24"
output:
  word_document: default
  html_document: default
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggpubr)
#install.packages("mvtnorm")

library(inti)
library(lme4)
library(GAPIT)

```



```{r}

#setwd("C:\\Users\\naresh89\\Desktop\\My projects\\GWAS\\Analysis_2024\\BLINK_MLM")
myY <- read.table("C:\\Users\\naresh89\\Desktop\\My projects\\gwas\\seedlingvigor&seedweight\\GWAS_2026\\gwas_26\\blues.csv _phenotype.txt", head = TRUE)
myG <- read.table("C:\\Users\\naresh89\\Desktop\\My projects\\gwas\\seedlingvigor&seedweight\\GWAS_2026\\gwas_26\\blues.csv _genotype.hmp.txt", comment.char = "")

#myG <- read.table("mdp_genotype_test.hmp.txt", head = FALSE)
myGAPIT <- GAPIT(G=myG, output.numerical=TRUE)
myGD= myGAPIT$GD
myGM= myGAPIT$GM


#myGD=read.table("C:\\Users\\naresh89\\Desktop\\My projects\\GWAS\\Analysis_2024\\2023\\GAPIT.Genotype.Numerical.txt",head=T)
#myGM=read.table("C:\\Users\\naresh89\\Desktop\\My projects\\GWAS\\Analysis_2024\\2023\\GAPIT.Genotype.map.txt",head=T)
#Step 2: Run GAPIT
#setwd("C:\\Users\\naresh89\\Desktop\\My projects\\GWAS\\Analysis_2024\\2023\\BLINK_farmCPU")
myGAPIT <- GAPIT(
  Y=myY,
  GD=myGD,
  GM=myGM,
  PCA.total=3,
  SNP.MAF = 0.05,
  Multiple_analysis=TRUE,
  model=c("Blink", "farmCPU", "MLM")
  )

```
# ==============================================================================
# SCRIPT: EXTRACT SIGNIFICANT GWAS HITS (M_eff Corrected)
# ==============================================================================
library(data.table)

# 1. SETTINGS
# Path to your GWAS results
setwd(r"(C:\Users\naresh89\Desktop\My projects\gwas\seedlingvigor&seedweight\GWAS_2026\gwas_26\post_GWAS)")

# Threshold Calculation (Based on your PLINK analysis)
M_eff     <- 2513
threshold <- 0.05 / M_eff

print(paste("Using M_eff:", M_eff))
print(paste("Significance Threshold (P-value):", formatC(threshold, format="e", digits=2)))
print(paste("Significance Threshold (-log10): ", round(-log10(threshold), 2)))

# 2. FIND FILES
# Look for all GAPIT result files
files <- list.files(pattern = "GAPIT.Association.GWAS_Results.*\\.csv$")

print(paste("Found", length(files), "result files. Processing..."))

# 3. LOOP AND FILTER
all_significant_hits <- list()

for (f in files) {
  # Read file
  # fill=TRUE handles cases where some rows might have issues, though GAPIT is usually clean
  df <- fread(f)
  
  # Filter for Significant SNPs
  # We use 'P.value' column (GAPIT standard)
  sig_rows <- df[P.value < threshold]
  
  # If we found hits, process them
  if (nrow(sig_rows) > 0) {
    
    # --- Parse Filename to get Model & Trait ---
    # Filename format example: GAPIT.Association.GWAS_Results.BLINK.Aspect_Ratio.csv
    
    # Remove the standard prefix and suffix
    clean_name <- sub("GAPIT.Association.GWAS_Results.", "", f)
    clean_name <- sub(".csv", "", clean_name)
    
    # Identify Model (Check against known GAPIT model names)
    model <- "Unknown"
    if (grepl("BLINK", clean_name))        model <- "BLINK"
    else if (grepl("FarmCPU", clean_name)) model <- "FarmCPU"
    else if (grepl("MLM", clean_name))     model <- "MLM"
    else if (grepl("GLM", clean_name))     model <- "GLM"
    else if (grepl("CMLM", clean_name))    model <- "CMLM"
    else if (grepl("SUPER", clean_name))   model <- "SUPER"
    
    # Identify Trait
    # Logic: The trait is whatever is left in the name after removing the Model
    # We strip the model name and any leading/trailing dots
    trait_raw <- sub(model, "", clean_name)
    trait <- gsub("^\\.|\\.$", "", trait_raw) # Remove dot at start or end
    
    # Add Info Columns
    sig_rows$Model <- model
    sig_rows$Trait <- trait
    sig_rows$Source_File <- f
    
    # Add to list
    all_significant_hits[[f]] <- sig_rows
  }
}

# 4. COMBINE AND SAVE
if (length(all_significant_hits) > 0) {
  
  final_table <- rbindlist(all_significant_hits)
  
  # Sort by Trait, then Chromosome, then Position
  final_table <- final_table[order(Trait, Chr, Pos)]
  
  # Save to CSV
  output_filename <- "Significant_GWAS_Hits_ALL_Traits.csv"
  fwrite(final_table, output_filename)
  
  print("✅ SUCCESS!")
  print(paste("Extracted", nrow(final_table), "significant SNPs."))
  print(paste("Saved results to:", output_filename))
  
  # Preview
  print(head(final_table[, .(Trait, Model, SNP, Chr, Pos, P.value)], 10))
  
} else {
  print("⚠️ No significant SNPs found in any file using this threshold.")
}
#####################################
# ------------------------------------------------------------------------------
# MANHATTAN + QQ (With Lambda & Model Names)
# ------------------------------------------------------------------------------

library(tidyverse)
library(ggrepel)
library(patchwork)

# ==============================================================================
# 1. SETTINGS
# ==============================================================================
# Point to your results folder
res_dir <- setwd(r"(C:\Users\naresh89\Desktop\My projects\gwas\seedlingvigor&seedweight\GWAS_2026\gwas_26\post_GWAS)")
# --- THRESHOLDS ---
M_eff       <- 2513
thr_sig_p    <- 0.05 / M_eff
thr_sig_logp <- -log10(thr_sig_p)

print(paste("Significant Cutoff (Red Line):", round(thr_sig_logp, 2)))

# ==============================================================================
# 2. READ DATA
# ==============================================================================
files <- list.files(
  path = res_dir,
  pattern = "^GAPIT\\.Association\\.GWAS_Results\\..+\\.csv$",
  full.names = TRUE
)

if(length(files) == 0) stop("No GAPIT results found! Check the folder path.")

models_set <- c("BLINK", "FarmCPU")

parse_name <- function(path) {
  base  <- basename(path)
  clean <- sub("GAPIT\\.Association\\.GWAS_Results\\.", "", base)
  clean <- sub("\\.csv$", "", clean)
  
  parts <- strsplit(clean, "\\.")[[1]]
  mdl   <- intersect(parts, models_set)[1]
  
  if(is.na(mdl)) return(list(model="Unknown", trait=clean))
  
  trt   <- paste(parts[parts != mdl], collapse=".")
  list(model = mdl, trait = trt)
}

print("Loading data...")
gwas_all <- map_df(files, function(f){
  nm <- parse_name(f)
  df <- read.csv(f, header = TRUE, stringsAsFactors = FALSE)
  if(!"P.value" %in% names(df)) return(NULL)
  
  df %>%
    rename(SNP = SNP, Chr = Chr, Pos = Pos, P = P.value) %>%
    mutate(model = nm$model, trait = nm$trait) %>%
    select(SNP, Chr, Pos, P, model, trait)
})

# ==============================================================================
# 3. PLOTTING FUNCTIONS
# ==============================================================================

# Colors
model_cols <- c("FarmCPU"="#D55E00","BLINK"="#0072B2")

# Trait Names
nice_names <- c(
  "SVG_score"    = "Vigor Score",
  "canopy_coverage"     = "Canopy Coverage",
  "canopy_height" = "Canopy Height",
  "canopy_volume_index" = "Canopy Volume Index"
  )

prepare_manhattan <- function(df_trait){
  df <- df_trait %>% filter(!is.na(P), !is.na(Chr), !is.na(Pos))
  chr_info <- df %>%
    group_by(Chr) %>%
    summarise(chr_len = max(Pos), .groups="drop") %>%
    arrange(as.numeric(Chr)) %>%
    mutate(tot = cumsum(chr_len) - chr_len)
  
  df2 <- df %>%
    left_join(chr_info, by="Chr") %>%
    mutate(pos_cum = Pos + tot, logp = -log10(pmax(P, 1e-300))) %>%
    arrange(as.numeric(Chr), Pos)
  
  axis_df <- chr_info %>% mutate(center = tot + chr_len/2)
  list(data = df2, axis = axis_df)
}

make_manhattan_plot <- function(df_tr, sig_line, nice_names, model_cols){
  man <- prepare_manhattan(df_tr)
  dat <- man$data; axis_df <- man$axis
  tr_raw <- unique(dat$trait)[1]
  plot_title <- ifelse(tr_raw %in% names(nice_names), nice_names[tr_raw], tr_raw)
  
  ggplot(dat, aes(pos_cum, logp, colour = model)) +
    geom_point(size = 0.8, alpha = 0.7) +
    
    # Red Significant Line
    geom_hline(yintercept = sig_line, colour = "red", linetype = "dashed", linewidth = 0.5) +
    
    # Annotate Top Hits
    geom_text_repel(
      data = dat %>% filter(logp >= sig_line),
      aes(label = SNP), size = 2, max.overlaps = 20, colour = "black", show.legend=FALSE
    ) +
    
    scale_colour_manual(values = model_cols) +
    scale_x_continuous(breaks = axis_df$center, labels = axis_df$Chr) +
    labs(title = plot_title, x = "Chromosome", y = "-log10(P)") +
    theme_bw(base_size = 11) +
    theme(
      legend.position = "none",
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()
    )
}

make_qq_plot <- function(df_tr, model_cols){
  # Prepare QQ Data
  prep_qq <- function(pvec){
    pvec <- sort(pvec)
    n <- length(pvec)
    tibble(exp = -log10((1:n)/(n+1)), obs = -log10(pmax(pvec, 1e-300)))
  }
  
  qq_df <- df_tr %>%
    filter(!is.na(P)) %>%
    group_by(model) %>%
    summarise(data = list(prep_qq(P)), .groups = "drop") %>%
    unnest(cols = data)
  
  # --- ANNOTATION: Calculate Lambda and Create Label ---
  annot_df <- df_tr %>%
    filter(!is.na(P)) %>%
    group_by(model) %>%
    summarise(
      # Lambda Calculation: Median(Observed) / Median(Expected)
      lambda_val = median(qchisq(1 - P, 1)) / qchisq(0.5, 1),
      .groups = "drop"
    ) %>%
    arrange(model) %>%
    mutate(
      # Create label: "Model (λ = 1.05)"
      label_text = paste0(model, " (λ = ", sprintf("%.2f", lambda_val), ")"),
      
      # Positioning
      x_pos = 0.5, 
      max_y = max(qq_df$obs, na.rm=TRUE),
      y_pos = max_y * 0.9 - (row_number() - 1) * (max_y * 0.15)
    )
  
  ggplot(qq_df, aes(exp, obs, colour = model)) +
    geom_abline(slope = 1, intercept = 0, colour = "gray40") +
    geom_point(size = 0.8) +
    
    # --- COMBINED LABELS (Model + Lambda) ---
    geom_text(
      data = annot_df,
      aes(x = x_pos, y = y_pos, label = label_text, colour = model),
      hjust = 0, size = 3.2, fontface = "bold", show.legend = FALSE
    ) +
    
    scale_colour_manual(values = model_cols) +
    labs(x = "Expected", y = "Observed") +
    theme_bw(base_size = 11) +
    theme(legend.position = "none") 
}

# ==============================================================================
# 4. BUILD & SAVE
# ==============================================================================
out_dir <- file.path(res_dir, "plots_combined_v3")
dir.create(out_dir, showWarnings = FALSE)

all_traits <- unique(gwas_all$trait)
combined_list <- list()

print("Generating plots...")

for (tr in all_traits){
  df_tr <- gwas_all %>% filter(trait == tr)
  
  p1 <- make_manhattan_plot(df_tr, thr_sig_logp, nice_names, model_cols)
  p2 <- make_qq_plot(df_tr, model_cols)
  
  combined <- (p1 + p2) + plot_layout(ncol = 2, widths = c(3, 1))
  combined_list[[tr]] <- combined
  
  ggsave(file.path(out_dir, paste0("GWAS_", tr, ".png")), combined, width = 10, height = 4, dpi = 300)
}

print("Saving stacked summary...")
final_stack <- wrap_plots(combined_list, ncol = 1)
ggsave(file.path(out_dir, "GWAS_Summary_All_Traits_With_Lambda.png"), 
       final_stack, width = 10, height = 3.5 * length(all_traits), dpi = 300, limitsize = FALSE)

print(paste("✅ Done! Plots saved to:", out_dir))
###########################
# ==============================================================================
# SCRIPT: CALCULATE PVE, ALLELES, AND EFFECT SIZES
# ==============================================================================
library(data.table)

# 1. SETTINGS
#setwd("C:/Users/naresh89/Desktop/My projects/gwas/Seed image analysis/GWAS_Weilv9/GWAS_results")

hits_file  <- "Significant_GWAS_Hits_ALL_Traits.csv"
geno_file  <- "blues.csv _genotype.hmp.txt"  # HapMap format
pheno_file <- "blues.csv _phenotype.txt"

# 2. LOAD DATA
print("Loading files...")
hits  <- fread(hits_file)
pheno <- fread(pheno_file) # Auto-detects tab/comma
geno  <- fread(geno_file)

# 3. HELPER FUNCTION: CONVERT IUPAC TO NUMERIC (0, 1, 2)
# ------------------------------------------------------------------------------
# Returns a list with: $genotype (numeric vector), $major (char), $minor (char)
convert_genotype <- function(geno_vector) {
  # definitions
  iupac_map <- list(
    'A'=c('A','A'), 'C'=c('C','C'), 'G'=c('G','G'), 'T'=c('T','T'),
    'R'=c('A','G'), 'Y'=c('C','T'), 'S'=c('G','C'), 'W'=c('A','T'),
    'K'=c('G','T'), 'M'=c('A','C'), 'N'=c(NA,NA)
  )
  
  # Flatten to find Major/Minor alleles
  all_bases <- unlist(lapply(geno_vector, function(x) iupac_map[[x]]))
  all_bases <- all_bases[!is.na(all_bases)]
  
  if(length(all_bases) == 0) return(list(genotype=rep(NA, length(geno_vector)), major=NA, minor=NA))
  
  # Count Alleles
  counts <- sort(table(all_bases), decreasing=TRUE)
  major  <- names(counts)[1]
  minor  <- ifelse(length(counts) > 1, names(counts)[2], NA)
  
  # Encode: 0 = MajorHom, 1 = Het, 2 = MinorHom
  numeric_geno <- sapply(geno_vector, function(x) {
    bases <- iupac_map[[x]]
    if(any(is.na(bases))) return(NA)
    
    # Check match
    has_major <- major %in% bases
    has_minor <- !is.na(minor) && (minor %in% bases)
    
    if(has_major && !has_minor) return(0)      # Homozygous Major
    if(!has_major && has_minor) return(2)      # Homozygous Minor (or other)
    if(has_major && has_minor)  return(1)      # Heterozygous
    return(NA)
  })
  
  return(list(genotype=numeric_geno, major=major, minor=minor))
}

# 4. MAIN LOOP
# ------------------------------------------------------------------------------
print("Calculating Stats for Significant Hits...")

# Initialize new columns
hits$PVE <- NA
hits$Major_Allele <- NA
hits$Minor_Allele <- NA
hits$Calc_Effect  <- NA
hits$Calc_SE      <- NA

# Get Taxa names (Columns 12 onwards in HapMap usually)
# We check the column names of geno that exist in pheno
taxa_cols <- intersect(colnames(geno), pheno$Taxa)

for(i in 1:nrow(hits)) {
  snp_id <- hits$SNP[i]
  trait  <- hits$Trait[i]
  
  # 1. Get Genotype Data
  # Find row in geno file
  snp_row <- geno[geno$`rs#` == snp_id, ..taxa_cols]
  
  if(nrow(snp_row) == 0) next # SNP not found
  
  # Convert to vector
  geno_vec <- as.character(as.vector(snp_row[1,]))
  
  # Convert IUPAC -> Numeric
  geno_info <- convert_genotype(geno_vec)
  
  # 2. Get Phenotype Data
  pheno_vec <- pheno[match(taxa_cols, pheno$Taxa), ..trait][[1]]
  
  # 3. Create Data Frame for LM
  df <- data.frame(Y = pheno_vec, X = geno_info$genotype)
  df <- na.omit(df) # Remove missing data
  
  if(nrow(df) < 5) next # Skip if too few lines
  
  # 4. Run Linear Model
  model <- lm(Y ~ X, data = df)
  s     <- summary(model)
  
  # 5. Extract Stats
  # Effect of Minor Allele (Slope of X)
  beta <- coef(s)["X", "Estimate"]
  se   <- coef(s)["X", "Std. Error"]
  r2   <- s$r.squared
  
  # Store
  hits$PVE[i]          <- r2
  hits$Calc_Effect[i]  <- beta
  hits$Calc_SE[i]      <- se
  hits$Major_Allele[i] <- geno_info$major
  hits$Minor_Allele[i] <- geno_info$minor
}

# 5. SAVE
write.csv(hits, "Significant_GWAS_Hits_With_Stats.csv", row.names=FALSE)
print("✅ Done! Saved 'Significant_GWAS_Hits_With_Stats.csv'")
print(head(hits[, c("SNP", "Trait", "PVE", "Calc_Effect")]))

