---
title: "Spatial Analysis of Seedling Vigor Traits for BLUEs Generation"
subtitle: "Two-Stage Spatial Analysis Using SpATS for GWAS"
author: "Venkata Naresh Boddepalli"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    number_sections: true
    theme: flatly
    fig_width: 10
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", dpi = 150
)
```

# Setup and Data Loading

## Load Required Packages

```{r load-packages}
library(SpATS)
library(lme4)
library(lmerTest)
library(emmeans)
library(ggplot2)
library(dplyr)
library(tidyr)
library(reshape2)
library(corrplot)
library(moments)
library(MASS)
library(car)
library(dplyr)   # reload dplyr last to unmask MASS::select
library(patchwork)
library(viridis)
library(ggpubr)

# Set output directory
out_dir <- getwd()
```

## Read Data

```{r read-data}
raw <- read.csv("rawdata_vigor_all.csv", stringsAsFactors = FALSE)

# Convert to appropriate types
raw$env    <- as.factor(raw$env)
raw$block  <- as.factor(raw$block)
raw$line   <- as.factor(raw$line)
raw$row    <- as.numeric(raw$row)
raw$column <- as.numeric(raw$column)
raw$plot   <- as.numeric(raw$plot)

# Define trait columns
traits <- c(
  "manual_SVG_score", "canopy_height", "canopy_area", 
  "canopy_coverage", "canopy_volume_index", 
  "ExG", "ExGR", "NGRDI", "GLI", "GRVI", 
  "RGBVI", "MGRVI", "VARI", "GBVI", 
  "ExR", "ExB", "CIVE", "NGBDI"
)

cat("Dimensions:", dim(raw), "\n")
cat("Environments:", levels(raw$env), "\n")
cat("Genotypes:", nlevels(raw$line), "\n")
cat("Blocks:", levels(raw$block), "\n")
str(raw)
summary(raw[, traits])
```

# Exploratory Data Analysis

## Distribution of Traits (Overall)

```{r eda-histograms, fig.height=10, fig.width=12}
raw_long <- raw %>%
  pivot_longer(cols = all_of(traits), names_to = "trait", values_to = "value")

ggplot(raw_long, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "steelblue", alpha = 0.7) +
  geom_density(color = "red", linewidth = 0.8) +
  facet_wrap(~ trait, scales = "free", ncol = 3) +
  theme_bw() +
  labs(title = "Distribution of Raw Trait Values (All Environments)")
```

## Boxplots per Environment

```{r eda-boxplots, fig.height=12, fig.width=14}
ggplot(raw_long, aes(x = env, y = value, fill = env)) +
  geom_boxplot(outlier.shape = 1, outlier.alpha = 0.4) +
  facet_wrap(~ trait, scales = "free_y", ncol = 3) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(title = "Trait Distributions by Environment") +
  scale_fill_viridis_d()
```

## Correlation Among Traits

```{r eda-correlation, fig.height=7, fig.width=8}
cor_mat <- cor(raw[, traits], use = "complete.obs")
corrplot(cor_mat, method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.8,
         tl.col = "black", tl.cex = 0.9,
         title = "Trait Correlations (Raw Data)", mar = c(0,0,2,0))
```

## Spatial Heatmaps of Raw Data

```{r eda-spatial-heatmaps, fig.height=16, fig.width=14}
for (tr in traits) {
  p_list <- list()
  for (e in levels(raw$env)) {
    d_sub <- raw[raw$env == e, ]
    p_list[[e]] <- ggplot(d_sub, aes(x = column, y = row, fill = .data[[tr]])) +
      geom_tile() +
      scale_fill_viridis_c(option = "C") +
      theme_minimal() +
      labs(title = e, fill = tr) +
      coord_equal()
  }
  combined <- wrap_plots(p_list, ncol = 3) +
    plot_annotation(title = paste("Spatial Heatmap:", tr))
  print(combined)
}
```

# Outlier Detection and Removal

Outliers are detected using model-based standardized residuals from an initial SpATS fit.
Observations with |standardized residual| > 3 are flagged and removed iteratively.

```{r outlier-detection}
# Helper: compute nseg as ~1/3 of unique positions, clamped to [4, 20]
get_nseg <- function(d) {
  nc <- max(4, min(20, length(unique(d$column)) %/% 3))
  nr <- max(4, min(20, length(unique(d$row))    %/% 3))
  c(nc, nr)
}

# Function to fit SpATS and detect outliers
# genotype as random for speed during outlier detection
fit_spats_detect_outliers <- function(data, trait, env_name, threshold = 3) {
  d <- data[data$env == env_name, ]
  d$geno <- as.factor(as.character(d$line))
  nseg <- get_nseg(d)

  tryCatch({
    fit <- SpATS(
      response  = trait,
      genotype  = "geno",
      genotype.as.random = TRUE,
      spatial   = ~ PSANOVA(column, row, nseg = nseg, degree = 3, pord = 2),
      fixed     = ~ block,
      data      = d,
      control   = list(tolerance = 1e-03, monitoring = 0)
    )

    resid_std <- residuals(fit) / sd(residuals(fit), na.rm = TRUE)
    outlier_idx <- which(abs(resid_std) > threshold)

    return(list(
      fit = fit,
      outliers = outlier_idx,
      n_outliers = length(outlier_idx),
      resid_std = resid_std
    ))
  }, error = function(e) {
    message(paste("SpATS error for", trait, "in", env_name, ":", e$message))
    return(NULL)
  })
}

# Apply outlier detection across all envs and traits
outlier_summary <- data.frame()
dat_clean <- raw  # working copy

for (tr in traits) {
  for (e in levels(raw$env)) {

    # Iterative outlier removal (max 2 rounds)
    for (iter in 1:2) {
      result <- fit_spats_detect_outliers(dat_clean, tr, e, threshold = 3)

      if (is.null(result) || result$n_outliers == 0) break

      # Get the row indices in dat_clean
      env_rows <- which(dat_clean$env == e)
      remove_rows <- env_rows[result$outliers]

      outlier_summary <- rbind(outlier_summary, data.frame(
        trait = tr, env = e, iteration = iter,
        n_outliers = result$n_outliers
      ))

      # Set outlier values to NA (rather than removing rows, to preserve structure)
      dat_clean[[tr]][remove_rows] <- NA
    }
  }
}

cat("\n=== Outlier Removal Summary ===\n")
if (nrow(outlier_summary) > 0) {
  print(outlier_summary)
  cat("\nTotal NAs introduced per trait:\n")
  print(colSums(is.na(dat_clean[, traits])))
} else {
  cat("No outliers detected.\n")
}
```

# Normality and Assumption Checks

After outlier removal, we verify the distributional assumptions by fitting SpATS models
on the cleaned data and examining residuals.

```{r normality-checks, fig.height=14, fig.width=12}
normality_results <- data.frame()

for (tr in traits) {
  qq_plots <- list()
  rv_plots <- list()

  for (e in levels(dat_clean$env)) {
    d <- dat_clean[dat_clean$env == e & !is.na(dat_clean[[tr]]), ]
    d$geno <- as.factor(as.character(d$line))
    nseg <- get_nseg(d)

    fit <- tryCatch({
      SpATS(
        response  = tr,
        genotype  = "geno",
        genotype.as.random = TRUE,
        spatial   = ~ PSANOVA(column, row, nseg = nseg, degree = 3, pord = 2),
        fixed     = ~ block,
        data      = d,
        control   = list(tolerance = 1e-03, monitoring = 0)
      )
    }, error = function(err) NULL)

    if (is.null(fit)) next

    res <- residuals(fit)
    fitted_vals <- fitted(fit)

    # Shapiro-Wilk on subsample (max 5000)
    set.seed(42)
    res_sample <- if (length(res) > 5000) sample(res, 5000) else res
    sw <- shapiro.test(res_sample)

    skew_val <- skewness(res, na.rm = TRUE)
    kurt_val <- kurtosis(res, na.rm = TRUE)

    normality_results <- rbind(normality_results, data.frame(
      trait = tr, env = e,
      shapiro_W = round(sw$statistic, 4),
      shapiro_p = format.pval(sw$p.value, digits = 3),
      skewness = round(skew_val, 3),
      kurtosis = round(kurt_val, 3)
    ))

    # Q-Q plot
    qq_plots[[e]] <- ggplot(data.frame(res = res), aes(sample = res)) +
      stat_qq(size = 0.8, alpha = 0.5) +
      stat_qq_line(color = "red") +
      theme_bw() +
      labs(title = e)

    # Residual vs fitted
    rv_plots[[e]] <- ggplot(data.frame(fitted = fitted_vals, res = res),
                            aes(x = fitted, y = res)) +
      geom_point(size = 0.6, alpha = 0.4) +
      geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
      geom_smooth(method = "loess", se = FALSE, color = "blue", linewidth = 0.5) +
      theme_bw() +
      labs(title = e)
  }

  # Print Q-Q plots
  if (length(qq_plots) > 0) {
    combined_qq <- wrap_plots(qq_plots, ncol = 3) +
      plot_annotation(title = paste("Q-Q Plots of Residuals:", tr))
    print(combined_qq)
  }

  # Print Residual vs Fitted plots
  if (length(rv_plots) > 0) {
    combined_rv <- wrap_plots(rv_plots, ncol = 3) +
      plot_annotation(title = paste("Residuals vs Fitted:", tr))
    print(combined_rv)
  }
}

cat("\n=== Normality Test Results ===\n")
print(normality_results, row.names = FALSE)
# Save the summary table to your working directory
write.csv(normality_results, 
          file = file.path(out_dir, "Normality_Test_Results_Summary.csv"), 
          row.names = FALSE)
```

# Environment-wise Spatial Analysis (SpATS) and BLUEs

Using SpATS with genotype as **fixed** effect to extract BLUEs.
Model: `trait ~ genotype(fixed) + block(fixed) + PSANOVA(column, row)`

```{r env-wise-blues, fig.height=6, fig.width=10}
blues_env_list <- list()
spats_diagnostics <- data.frame()

for (tr in traits) {
  cat("\n========================================\n")
  cat("Processing trait:", tr, "\n")
  cat("========================================\n")

  for (e in levels(dat_clean$env)) {
    d <- dat_clean[dat_clean$env == e & !is.na(dat_clean[[tr]]), ]
    d$geno <- as.factor(as.character(d$line))
    nseg <- get_nseg(d)

    # Genotype as FIXED for BLUEs extraction
    fit <- tryCatch({
      SpATS(
        response  = tr,
        genotype  = "geno",
        genotype.as.random = FALSE,
        spatial   = ~ PSANOVA(column, row, nseg = nseg, degree = 3, pord = 2),
        fixed     = ~ block,
        data      = d,
        control   = list(tolerance = 1e-03, monitoring = 0)
      )
    }, error = function(err) {
      message(paste("SpATS error:", tr, "in", e, "-", err$message))
      NULL
    })

    if (is.null(fit)) next

    # Extract BLUEs (genotype coefficients from fixed effects)
    blues <- predict(fit, which = "geno")
    blues_df <- data.frame(
      line  = blues$geno,
      env   = e,
      trait = tr,
      BLUE  = blues$predicted.values,
      SE    = blues$standard.errors
    )
    blues_env_list[[paste(tr, e, sep = "_")]] <- blues_df

    # Effective dimensions (spatial complexity)
    ed <- fit$eff.dim
    spats_diagnostics <- rbind(spats_diagnostics, data.frame(
      trait = tr, env = e,
      ED_f_col     = round(ed[["f(column)"]],        2),
      ED_f_row     = round(ed[["f(row)"]],           2),
      ED_f_col_row = round(ed[["f(column):f(row)"]], 2),
      deviance     = round(fit$deviance, 2)
    ))

    # Plot spatial trend
    cat(paste("\nSpatial trend plot:", tr, "-", e, "\n"))
    tryCatch(plot(fit, main = paste(tr, "-", e)), error = function(err) NULL)
  }
}

cat("\n=== SpATS Effective Dimensions ===\n")
print(spats_diagnostics, row.names = FALSE)
```

## Compile Environment-wise BLUEs

```{r compile-env-blues}
blues_env_all <- do.call(rbind, blues_env_list)
rownames(blues_env_all) <- NULL

# Summary
cat("Environment-wise BLUEs summary:\n")
cat("Total records:", nrow(blues_env_all), "\n")
cat("Traits:", unique(blues_env_all$trait), "\n")
cat("Environments:", unique(blues_env_all$env), "\n")
cat("Genotypes:", length(unique(blues_env_all$line)), "\n")

# Reshape to wide format for output
blues_env_wide <- blues_env_all %>%
  dplyr::select(line, env, trait, BLUE) %>%
  pivot_wider(names_from = trait, values_from = BLUE)

head(blues_env_wide)

# Save environment-wise BLUEs
write.csv(blues_env_wide, file.path(out_dir, "BLUEs_per_environment.csv"),
          row.names = FALSE)
cat("\nSaved: BLUEs_per_environment.csv\n")
```

## Visualize Environment-wise BLUEs

```{r viz-env-blues, fig.height=10, fig.width=12}
ggplot(blues_env_all, aes(x = env, y = BLUE, fill = env)) +
  geom_boxplot(outlier.shape = 1, outlier.alpha = 0.3) +
  facet_wrap(~ trait, scales = "free_y", ncol = 3) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(title = "",
       y = "BLUE Value") +
  scale_fill_viridis_d()
# First, run your plot code
p_env_blues <- ggplot(blues_env_all, aes(x = env, y = BLUE, fill = env)) +
  geom_boxplot(outlier.shape = 1, outlier.alpha = 0.3) +
  facet_wrap(~ trait, scales = "free_y", ncol = 3) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(title = "", y = "BLUE Value") +
  scale_fill_viridis_d()

# Then, save it with specific dimensions for 18 facets
ggsave(filename = "Boxplots_Env_Wise_BLUEs.png", 
       plot = p_env_blues,
       width = 15,    # Wide enough for 3 columns
       height = 20,   # Tall enough for 6 rows of facets
       units = "in", 
       dpi = 300)     # High resolution for manuscript
```

# Combined BLUEs Across All Environments (Two-Stage)

Stage 2: Using environment-wise BLUEs as input, fit a mixed model with
genotype as **fixed** and environment as **random** to obtain combined BLUEs.

Model: `BLUE_ij ~ line_i (fixed) + (1|env_j)`

```{r combined-blues}
combined_blues_list <- list()
heritability_results <- data.frame()

for (tr in traits) {
  cat("\n--- Combined analysis for:", tr, "---\n")

  # Subset env-wise BLUEs for this trait
  d_tr <- blues_env_all[blues_env_all$trait == tr, ]
  d_tr$line <- as.factor(d_tr$line)
  d_tr$env  <- as.factor(d_tr$env)

  n_envs <- nlevels(d_tr$env)

  # Fit mixed model: line (fixed) + env (random)
  fit_combined <- tryCatch({
    lmer(BLUE ~ line + (1 | env), data = d_tr)
  }, error = function(err) {
    message(paste("lmer error for", tr, ":", err$message))
    NULL
  })

  if (is.null(fit_combined)) next

  # Extract combined BLUEs using emmeans
  emm <- emmeans(fit_combined, ~ line)
  emm_df <- as.data.frame(emm)

  combined_blues_list[[tr]] <- data.frame(
    line  = emm_df$line,
    trait = tr,
    BLUE  = emm_df$emmean,
    SE    = emm_df$SE
  )

  # Variance components for heritability
  vc <- as.data.frame(VarCorr(fit_combined))
  sigma2_e <- vc$vcov[vc$grp == "Residual"]
  sigma2_env <- vc$vcov[vc$grp == "env"]

  # Also fit with genotype as random for heritability estimation
  fit_h2 <- tryCatch({
    lmer(BLUE ~ (1 | line) + (1 | env), data = d_tr)
  }, error = function(err) NULL)

  if (!is.null(fit_h2)) {
    vc_h2 <- as.data.frame(VarCorr(fit_h2))
    sigma2_g   <- vc_h2$vcov[vc_h2$grp == "line"]
    sigma2_res <- vc_h2$vcov[vc_h2$grp == "Residual"]
    sigma2_env_h2 <- vc_h2$vcov[vc_h2$grp == "env"]

    # Broad-sense heritability (entry-mean basis)
    # H2 = sigma2_g / (sigma2_g + sigma2_res/n_envs)
    H2 <- sigma2_g / (sigma2_g + sigma2_res / n_envs)

    heritability_results <- rbind(heritability_results, data.frame(
      trait = tr,
      sigma2_geno = round(sigma2_g, 6),
      sigma2_env  = round(sigma2_env_h2, 6),
      sigma2_resid = round(sigma2_res, 6),
      H2_entry_mean = round(H2, 4),
      n_envs = n_envs
    ))

    cat(paste("  H² (entry-mean basis):", round(H2, 4), "\n"))
  }
}
```

## Compile and Save Combined BLUEs

```{r save-combined-blues}
combined_blues_all <- do.call(rbind, combined_blues_list)
rownames(combined_blues_all) <- NULL

# Reshape to wide format
combined_blues_wide <- combined_blues_all %>%
  dplyr::select(line, trait, BLUE) %>%
  pivot_wider(names_from = trait, values_from = BLUE)

cat("Combined BLUEs summary:\n")
cat("Genotypes:", nrow(combined_blues_wide), "\n")
cat("Traits:", ncol(combined_blues_wide) - 1, "\n")
head(combined_blues_wide)

write.csv(combined_blues_wide, file.path(out_dir, "BLUEs_combined.csv"),
          row.names = FALSE)
cat("\nSaved: BLUEs_combined.csv\n")
```

## Heritability Estimates

```{r heritability}
cat("\n=== Broad-Sense Heritability (Entry-Mean Basis) ===\n")
print(heritability_results, row.names = FALSE)
# 1. Save the raw heritability results as a CSV
write.csv(heritability_results, 
          file = file.path(out_dir, "Heritability_Results_All_Traits.csv"), 
          row.names = FALSE)

# 2. Create a "Manuscript-Ready" formatted version
# This rounds the values and renames columns for professional presentation
heritability_formatted <- heritability_results %>%
  mutate(
    H2_entry_mean = round(H2_entry_mean, 3),
    sigma2_geno = round(sigma2_geno, 5),
    sigma2_env = round(sigma2_env, 5),
    sigma2_resid = round(sigma2_resid, 5)
  ) %>%
  rename(
    Trait = trait,
    `Genetic Variance` = sigma2_geno,
    `Environmental Variance` = sigma2_env,
    `Residual Variance` = sigma2_resid,
    `Heritability (H2)` = H2_entry_mean,
    `Environments (n)` = n_envs
  )

# Save the formatted version
write.csv(heritability_formatted, 
          file = file.path(out_dir, "Table_Heritability_Formatted.csv"), 
          row.names = FALSE)

cat("\nSaved: Heritability results have been exported to your working directory.\n")
```

## Visualize Combined BLUEs

```{r viz-combined, fig.height=10, fig.width=12}
# Histograms of combined BLUEs
combined_long <- combined_blues_all %>%
  dplyr::select(line, trait, BLUE)

ggplot(combined_long, aes(x = BLUE)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "steelblue", alpha = 0.7) +
  geom_density(color = "red", linewidth = 0.8) +
  facet_wrap(~ trait, scales = "free", ncol = 3) +
  theme_bw() +
  labs(title = "Distribution of Combined BLUEs Across 6 Environments")
```

```{r combined-correlations, fig.height=7, fig.width=8}
# Correlation among combined BLUEs
cor_combined <- cor(combined_blues_wide[, -1], use = "complete.obs")
corrplot(cor_combined, method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.8,
         tl.col = "black", tl.cex = 0.9,
         title = "", mar = c(0,0,2,0))
# Define the file path and settings
# Using a high resolution (res = 300) for manuscript quality
png(file.path(out_dir, "Figure_Trait_Correlations_BLUEs.png"), 
    width = 10, 
    height = 10, 
    units = "in", 
    res = 300)

# Run the correlation plot command
corrplot(cor_combined, 
         method = "color", 
         type = "upper",
         addCoef.col = "black", 
         number.cex = 0.7,      # Reduced slightly to ensure numbers fit
         tl.col = "black", 
         tl.cex = 0.8,          # Adjusted for 18 traits
         title = "", 
         mar = c(0,0,2,0))

# Close the device to save the file
dev.off()

cat("\nSuccess: Correlation plot saved to", out_dir, "\n")
```

# GxE Visualization

```{r gxe-viz, fig.height=10, fig.width=12}
# Check genotype rank changes across environments for key traits
for (tr in c("SVG_score", "canopy_height", "canopy_coverage")) {
  d_tr <- blues_env_all[blues_env_all$trait == tr, ]

  # Select top and bottom 10 genotypes based on mean BLUE
  geno_means <- d_tr %>%
    group_by(line) %>%
    summarize(mean_blue = mean(BLUE, na.rm = TRUE)) %>%
    arrange(desc(mean_blue))

  top_bottom <- c(head(geno_means$line, 10), tail(geno_means$line, 10))

  d_sub <- d_tr[d_tr$line %in% top_bottom, ]

  p <- ggplot(d_sub, aes(x = env, y = BLUE, group = line, color = line)) +
    geom_line(alpha = 0.7) +
    geom_point(size = 1.5) +
    theme_bw() +
    theme(legend.position = "right",
          legend.text = element_text(size = 6),
          axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = paste("GxE Interaction:", tr, "(Top & Bottom 10 Genotypes)"),
         y = "BLUE Value", color = "Genotype")

  print(p)
}
```

# Session Info

```{r session-info}
sessionInfo()
```
